Ion pairing controls rheological properties of polyelectrolyte crocodile line hydrogels

### 2016-01-12 ion_pairing_paper.tex
Dispatch former discussion into results to introduce properly HSAB theory upfront.

\subsection{Water swelling and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer compositions. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 8\%wt suspension as a biphasic mixture of its individual components.

Therefore, we are able to dissolve the five polymers that have the weakest ion pairing, but neither the two fluorides nor \ce{PIm+Cl-} where counterions are expected to be strongly bounded to the chain. This means that in absence of charge dissociation the chain is in poor solvent. Furthermore, we held the two fluoride polymers in water at boiling temperature overnight without any dissolution, implying that the $\Theta$ temperature is higher than \SI{100}{\celsius}. Nonetheless, due to charge dissociation, most of our polymers are soluble at \SI{25}{\celsius}.

@@

In Figure~\ref{fig:strain}a we compare the rheological responses of Poly(pyrrolidinium) based hydrogels for the three counterions allowing dissolution in water. Gels with bromide and iodide counterions show roughly the same mechanical behavior in the linear regime but \ce{PPyr+I-} yields at slightly higher strains ($\gamma_c=4\%$ instead of $3\%$). By contrast, the gel with chloride counterions is weaker by more than one order of magnitude.

We also compare the rheology of \ce{PIm+X-} and \ce{PPyr+X-} with the same counterion \ce{X} in Figure~\ref{fig:strain}b,c. For both iodide (Figure~\ref{fig:strain}b, X=I) and bromide (Figure~\ref{fig:strain}c, X=Br), imidazolium based gels are much weaker than their pyrrolidinium counterpart.

@@

\subsection{Discussion}

The above dependence of the gel parameters on the nature of the counterions could be rationalized on the basis of ion-pairing processes. The unsolubility/unswellability of \ce{PIm+F-}, \ce{PPyr+F-} and \ce{PIm+Cl-} suggest that ion pairs are not efficiently dissociated in aqueous medium, thus hindering the solvation process. By comparison, \ce{PIm+Br-}, \ce{PIm+I-}, \ce{PPyr+Br-} and \ce{PPyr+I-} form homogeneous materials when suspended in water, which one can correlate with efficient dissociation of the halide counterions from the polymer main chain is fully consistency with decreased ionic bond strength. Such dissociation, in turns makes it possible for the ``hard'' phosphonate group on each polymer chain's end to interact with the ``free'' cationic groups of the side chains, which we showed in a previous paper to be at the origin of the gel formation mechanism\cite{Srour2014}.

The similarity in gel parameters observed for \ce{PPyr+Br-} and \ce{PPyr+I-} ($G^\prime\approx\SI{8500}{\pascal}$) suggests that full ion pair dissociation is achieved in the latter, or at least that this dissociation is enough in both cases to provide maximal interaction between the terminal phosphonate and the pyrrolidinium side groups. Conversely the significantly different behavior between \ce{PIm+I-} ($G^\prime= \SI{78}{\pascal}$) and \ce{PIm+Br-} ($G^\prime= \SI{13}{\pascal}$) indicates that dissociation is only partial and occurs to a lesser extent in the latter than in the former. The same conclusion can be drawn for \ce{PPyr+Cl-}, in which the elastic modulus ($G^\prime= \SI{270}{\pascal}$) value drops by more than one order of magnitude as compared to its bromide and iodide counterparts.

As already mentioned (vide supra) another striking feature of these gels is that in similar conditions poly(pyrrolidinium) based gels are systematically much stronger than those formed with poly(imidazolium). Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge; consequently, according to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected for the latter with the ``soft'' halide counterions\cite{Goossens2009}. As a consequence, for a given halide, dissociation is less favored in the case of poly(imidazolium), as clearly illustrated by the difference in solvation between \ce{PIm+Cl-} and \ce{PPyr+Cl-}. Conversely, association of the lateral cations with the ``hard'' phosphonate moieties at the polymer chain’s end is expected to be stronger with pyrrolidinium, hence a systematically larger elastic modulus. The interplay of these two effects contributes in making poly(pyrrolidinium) based gels much stronger than their poly(imidazolium) counterparts; comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ value (13 and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude.

>>>>

\subsection{Water swelling, ion-pairing and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer compositions. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 8\%wt suspension as a biphasic mixture of its individual components.

The above dependence of gel formability on the nature of the counterions could be rationalized on the basis of ion-pairing processes. According to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected between a ``softer'' cationic moiety and a ``harder'' halide counterions\cite{Goossens2009}. Therefore charge dissociation in water and thus solubility of the polymer should be easier with larger halide and pyrrolidinium. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge. That is why we are able to dissolve the five polymers that have the weakest ion pairing, but neither the two fluorides nor \ce{PIm+Cl-} where counterions are expected to be strongly bounded to the chain.

A fortiori, the insolubility in absence of charge dissociation indicates that the chain is in poor solvent. To estimate $\Theta$ temperature, we held the two fluoride polymers in water at boiling temperature overnight without any dissolution, implying $\Theta>\SI{100}{\celsius}$. Nonetheless, due to charge dissociation, five of our polyelectrolytes are soluble at \SI{25}{\celsius}.

@@

In Figure~\ref{fig:strain}a we compare the rheological responses of Poly(pyrrolidinium) based hydrogels for the three counterions allowing dissolution in water. Gels with bromide and iodide counterions show roughly the same mechanical behavior in the linear regime but \ce{PPyr+I-} yields at slightly higher strains ($\gamma_c=4\%$ instead of $3\%$). By contrast, the gel with chloride counterions is weaker by more than one order of magnitude. In other words, a ``soft'' halide counterion, efficiently dissociated, is correlated to a harder gel.

We also compare the rheology of \ce{PIm+X-} and \ce{PPyr+X-} with the same counterion \ce{X} in Figure~\ref{fig:strain}b,c. For both iodide (Figure~\ref{fig:strain}b, X=I) and bromide (Figure~\ref{fig:strain}c, X=Br), imidazolium based gels are much weaker than their pyrrolidinium counterpart with a broader linear regime. Comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ value (13 and \SI{8100}{\pascal}, resp.) and $\gamma_c$ value (804\% and 1.6\%, resp.) differ by almost three orders of magnitude. Here also, gel hardness is correlated to efficient charge dissociation, this time via ``harder'' cationic moieties. In the following we explain quantitatively this correlation from a microscopic model.

@@

%\subsection{Discussion}
%
%The above dependence of the gel parameters on the nature of the counterions could be rationalized on the basis of ion-pairing processes. The unsolubility/unswellability of \ce{PIm+F-}, \ce{PPyr+F-} and \ce{PIm+Cl-} suggest that ion pairs are not efficiently dissociated in aqueous medium, thus hindering the solvation process. By comparison, \ce{PIm+Br-}, \ce{PIm+I-}, \ce{PPyr+Br-} and \ce{PPyr+I-} form homogeneous materials when suspended in water, which one can correlate with efficient dissociation of the halide counterions from the polymer main chain is fully consistency with decreased ionic bond strength. Such dissociation, in turns makes it possible for the ``hard'' phosphonate group on each polymer chain's end to interact with the ``free'' cationic groups of the side chains, which we showed in a previous paper to be at the origin of the gel formation mechanism\cite{Srour2014}.
%
%The similarity in gel parameters observed for \ce{PPyr+Br-} and \ce{PPyr+I-} ($G^\prime\approx\SI{8500}{\pascal}$) suggests that full ion pair dissociation is achieved in the latter, or at least that this dissociation is enough in both cases to provide maximal interaction between the terminal phosphonate and the pyrrolidinium side groups. Conversely the significantly different behavior between \ce{PIm+I-} ($G^\prime= \SI{78}{\pascal}$) and \ce{PIm+Br-} ($G^\prime= \SI{13}{\pascal}$) indicates that dissociation is only partial and occurs to a lesser extent in the latter than in the former. The same conclusion can be drawn for \ce{PPyr+Cl-}, in which the elastic modulus ($G^\prime= \SI{270}{\pascal}$) value drops by more than one order of magnitude as compared to its bromide and iodide counterparts.
%
%As already mentioned (vide supra) another striking feature of these gels is that in similar conditions poly(pyrrolidinium) based gels are systematically much stronger than those formed with poly(imidazolium). Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge; consequently, according to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected for the latter with the ``soft'' halide counterions\cite{Goossens2009}. As a consequence, for a given halide, dissociation is less favored in the case of poly(imidazolium), as clearly illustrated by the difference in solvation between \ce{PIm+Cl-} and \ce{PPyr+Cl-}. Conversely, association of the lateral cations with the ``hard'' phosphonate moieties at the polymer chain’s end is expected to be stronger with pyrrolidinium, hence a systematically larger elastic modulus. The interplay of these two effects contributes in making poly(pyrrolidinium) based gels much stronger than their poly(imidazolium) counterparts; comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ value (13 and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude.
